Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-07-11 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

\medskip
Viscous finger morphologies were investigated by \citet{guan_2003} based on mathematical formulation and semi-analytic solutions of the coupled Darcy and continuity equations developed by \citet{mclean_1981}. They conducted a sensitivity analysis to investigate fingers' formation, dimensions and branchiness for a range of viscosity ratios $\left(\right.$10$^{2}\le\textit{VR}\le$10$\left.^{4}\right)$ and modified capillary numbers $\left(\right.$632$\le N_{C}^{'}\le$6.32$\times$10$^{7}$, with $N_{C}^{'}= U_{f}\mu\gamma^{-1}(W/b)^{2}$, where $U_{f}$ is the velocity of the finger, $W$ is half Hele-Shaw cell width and $b$ is the cell thickness$\left.\right)$. Finger width $\left(\lambda_{f}\right)$ and volumetric flow rate $\left(\mathcal{Q}\right)$ at the outflow region were correlated through,   
\begin{equation}   
   \mathcal{Q} = U_{f} b \lambda_{f}.\label{eqn:guaneqn} 
\end{equation}

\noindent Here, numerical simulations performed with viscosity ratio of 10 and 150 indicated maximum fingers width of approximately 0.45-0.70 and 0.50-0.90 cm (Fig.~\ref{fig:homoheleshaw_VN10_VN150} a and b), respectively (assuming b = 0.1 cm). This is in close agreement with expected values obtained from Eqn.~\ref{eqn:guaneqn} which indicates maximum finger width ranging from 0.13 to 0.75 (for {\it VR}=10) and from 0.30 to 0.75 ({\it VR}=150, Table~\ref{table:Heleshaw}).

@@

\medskip
In order to qualitatively investigate the impact of heterogeneity (here represented by changing in the permeability field), numerical simulations were conducted using a prescribed permeability distribution and viscosity ratios of 10 and 150. Absolute permeability ranging from 1.0$\times$10$^{-12}$ to 5.0$\times$10$^{-10}$ cm$^{2}$ (\ie 0.1 $\le$ {\bf K} $\le$ 50 milidarcy) was used in the simulations, which were performed with the same geometry,  mesh resolution, boundary and initial conditions as in previous homogeneous cases(Figs.~\ref{fig:homoheleshaw_VN3}-\ref{fig:homoheleshaw_VN10_VN150}b).

\medskip
Figures~\ref{fig:HeleShawHeter_VR10} and \ref{fig:HeleShawHeter_VR150} show formation of multi-scale elongations with continuous growth and coalescence of dendritic finger branching in simulations performed with viscosity ratio of 10 and 150, respectively. In both simulations, maximum fingers width of approximately 0.44-0.56 ({\it VR}=10) and 0.24-0.44 cm were found, whereas estimated values (based on Eqn.~\ref{eqn:guaneqn}) were 0.25-4.50 ({\it VR}=10) and 0.88-3.00 cm ({\it VR}=150) -- Fig.~\ref{fig:homoheleshaw_VN10_VN150} c-d, Table~\ref{table:Heleshaw}. Such calculated ranges \citep[\ie theoretical values based on analytic solutions due to][]{mclean_1981} clearly overestimate fingers dimensions as the mathematical formulation (and therefore Eqn.~\ref{eqn:guaneqn}) assumes incompressible flows in homogeneous porous medium, \ie it does not take into account any spatial variability of heterogeneity.

>>>>

\medskip
Viscous finger morphologies were investigated by \citet{guan_2003} based on mathematical formulation and semi-analytic solutions of the coupled Darcy and continuity equations developed by \citet{mclean_1981}. They conducted a sensitivity analysis to investigate fingers' formation, dimensions and branchiness for a range of viscosity ratios $\left(\right.$10$^{2}\le\textit{VR}\le$10$\left.^{4}\right)$ and modified capillary numbers $\left(\right.$632$\le N_{C}^{'}\le$6.32$\times$10$^{7}$, with $N_{C}^{'}= U_{f}\mu\gamma^{-1}(W/b)^{2}$, where $U_{f}$ is the velocity of the finger, $W$ is half Hele-Shaw cell width and $b$ is the cell thickness$\left.\right)$. Finger width $\left(\lambda_{f}\right)$ and volumetric flow rate $\left(\mathcal{Q}\right)$ at the outflow region were correlated through,   
\begin{equation}   
   \mathcal{Q} = U_{f} b \lambda_{f}.\label{eqn:guaneqn} 
\end{equation}  
Here, numerical simulations performed with viscosity ratio of 10 and 150 indicated maximum fingers width of approximately 0.45-0.70 and 0.50-0.90 cm (Fig.~\ref{fig:homoheleshaw_VN10_VN150} a and b), respectively (assuming b = 0.1 cm). This is in close agreement with expected values obtained from Eqn.~\ref{eqn:guaneqn} which indicates maximum finger width ranging from 0.13 to 0.75 (for {\it VR}=10) and from 0.30 to 0.75 ({\it VR}=150, Table~\ref{table:Heleshaw}).

@@

In order to qualitatively investigate the impact of heterogeneity (here represented by changing in the permeability field), numerical simulations were conducted using a prescribed permeability distribution and viscosity ratios of 10 and 150. Absolute permeability ranging from 1.0$\times$10$^{-12}$ to 5.0$\times$10$^{-10}$ cm$^{2}$ (\ie 0.1 $\le$ {\bf K} $\le$ 50 milidarcy) was used in the simulations, which were performed with the same geometry,  mesh resolution, boundary and initial conditions as in previous homogeneous cases(Figs.~\ref{fig:homoheleshaw_VN3}-\ref{fig:homoheleshaw_VN10_VN150}b).

Figures~\ref{fig:HeleShawHeter_VR10} and \ref{fig:HeleShawHeter_VR150} show formation of multi-scale elongations with continuous growth and coalescence of dendritic finger branching in simulations performed with viscosity ratio of 10 and 150, respectively. In both simulations, maximum fingers width of approximately 0.44-0.56 ({\it VR}=10) and 0.24-0.44 cm were found, whereas estimated values (based on Eqn.~\ref{eqn:guaneqn}) were 0.25-4.50 ({\it VR}=10) and 0.88-3.00 cm ({\it VR}=150) -- Fig.~\ref{fig:homoheleshaw_VN10_VN150} c-d, Table~\ref{table:Heleshaw}. Such calculated ranges \citep[\ie theoretical values based on analytic solutions due to][]{mclean_1981} clearly overestimate fingers dimensions as the mathematical formulation (and therefore Eqn.~\ref{eqn:guaneqn}) assumes incompressible flows in homogeneous porous medium, \ie it does not take into account any spatial variability of heterogeneity.
